Regression Cube Analysis of Population Study Data

### 2015-06-22 draft.tex
replaced R² value appearances

%\subsection{Abstracting Regression Results Using $R^2$}
\subsection{Abstracting Regression Results}
\begin{figure}[htb]
 \centering
 \includegraphics[width=1.0\linewidth]{figures/cube}
 \caption{
 (a) Overview visualization using a mosaic plot for the formula $Z \sim X + Y$, where $Z$ assumes the feature $age$.
 %%
 The descriptive metrics extracted from the regression model are mapped to color saturation (for $R^2$, saturated color shows a strong correlation).
 %%
 (b) Now, $Z$ is set to all features $n$ and yields $n$ mosaic plot visualizations.
 %%
 These represent the slices in our cube.
 %%
 The metric describing the regression model of each slice voxel is mapped on opacity in the 3D view later on, reducing the occlusion of other values.
 }
  \label{fig:Cube}
\end{figure}
%%
\noindent The goal of an overview visualization is to provide a comprehensive view on the data (raw or using descriptive metrics \cite{Bertini}), which is easy to understand.
%%
As described in our previous work \cite{Klemm2014VIS}, correlation values scaled between 0 (no correlation) and 1 (perfect correlation) can be encoded with color in a mosaic plot.
%%
Regression models are more complex, having many associated describing metrics.
%%
For the \emph{Regression Cube} analysis we are interested in the quality-of-fit of the resulting model.
%%
This allows to infer about the predictive quality of the independent features included in the model.
%%
As described in Section~\ref{sec:RegressionAnalysis}, the $R^2$ \add{and adjusted $R^2$} value are metrics allowing for this kind of assessment.
%%
% \add{
% %%
% Users with in-depth knowledge about regression models are given the option to render the \emph{Regression Cube} using other descriptive features presented in \ref{sec:RegressionAnalysis}.
% %%
% Therefore, these experts can investigate the cube with emphasis on specific aspects of the model.
% %%
% To avoid shutting out users unfamiliar with these metrics, the \emph{Regression Cube} is set per default to show $R^2$ values.
% }
%%
\paragraph{2D (Slice) View.}
%%
Since $R^2$ is scaled between $[0, 1]$, it allows for comparison \emph{between} regression models.
%%
We can apply the same mosaic plot mapping by translating the $R^2$ values to \emph{color saturation} (Fig.~\ref{fig:Cube}a).
%%
This describes a 2D regression square for dynamic variables $X$ and $Y$ (e.g., $Age \sim X + Y$).
%%
\design{
%%
Based on expert feedback on early versions of this view, the amount of features used to compare regression models was extended.
%%
Therefore, these experts can investigate the cube with emphasis on specific aspects of the model.
%%
}
%%
\add{
%%
$Adjusted$ $R^2$ can be represented in the same way, since they share the scale.
%%
$AIC$ values are not restricted to a normalized scale.
%%
They have to be normalized in order to map it onto color saturation.
%%
%The resulting scale may be distort by outliers, which requires facilities for allowing the user to select scale ranges.
The resulting scale may be distorted by outliers derived from poor regression models.
%%
To tackle this problem, we provide a slider input, which maps the transfer function of the metric to color saturation based on user-selected ranges.
%%
Outliers can be cut off to emphasize ranges of interest.
%%
}
\design{
%%
To avoid shutting out users unfamiliar with these metrics, the \emph{Regression Cube} is set per default to show $R^2$ values.
%%
}

@@

\textbf{Show the Important} [\textbf{2.SI}]. The 3D cube visualization acts as an overview over the whole data set.
%%
Here, regression models with large $R^2$ values can be spotted fast, steering the user's attention to the respective slice.

@@

$Cancer \sim X + Y + Z$ subtracted with the $R^2$ from $Cancer \sim Age$ excludes the confounding effect that age has in view of the target $Cancer$ feature.
%%
This is achieved through cube comparison.
%%

\paragraph{Cube Comparison.}
\design{
%%
Cube comparisons were introduced later in the project.
%%
Epidemiologists with focus on statistics pointed out, that comparing outcomes of different formulas is suitable for removing the effect of possible confounding features.
%%
}
%%
\emph{Regression Cubes} can be compared by creating difference views.
%%
One cube (extracted with one formula) acts as reference.
%%
The absolute difference in $R^2$ values with the second cube is calculated, yielding a difference cube showing only the differences between the two formulas.
%%
For example, it can be utilized for comparing the influence of a single feature on the complete result (e.g., $Z \sim X + Y$ and $Z \sim X + Y + Income$).

@@

%%
\textbf{Difference Cube.}
Adding a formula also adds it to the reference selection for a difference cube.
%%
Since all cells in the cube are represented using $R^2$ values, a difference cube is calculated by denoting the absolute difference of $R^2$ for each cell.
%%
\subsection{Regression Cube Visualization [2.SI].}
%%
The visualization and interaction with the \emph{Regression Cube} is the prototype core.
%%
Results from the statistical processors are uploaded into the visualization slice by slice.
%%
This allows the assessment of the data as soon as parts of the calculations are finished while the rest is still in progress.

@@

\paragraph{3D Prism as Data Mini-Map.}
\design{
%%
In early prototype versions, the 3D prism acted as starting point for the data analysis without the implementation of a separate 2D view.
%%
Slices were shown using cutaway planes.
%%
This approach was not popular with our domain experts, because the complexity of the visualization overwhelmed them. %did not like the combination of 3D interaction and slicing with representatives.
%%
}
%%
The 3D cube representation \design{was redesigned to} act as an overview over the whole data set.
%%
Its purpose is not to derive detailed information about data points.
%%
It serves as a function similar to a mini-map, guiding the attention to points of interest in the data.
%%
It also gives context information about adjacent data values when using the 2D mosaic plot.
%%
\design{The distinction between overview and details and demand using two different representations was well received with our domain experts.}
%%
The displayed prism shows values above the matrix diagonal.
%%
For formulas with a dynamic target feature (e.g. exploratory analysis using $Z \sim X + Y$), the color encodes the absolute regression metric values (Fig.~\ref{fig:System}b).
%%
Applying this strategy to a formula containing a static target (e.g. $Cancer \sim X + Y + Z$) yields many occlusions, since the CFS algorithm creates the same feature space for every slice.
%%
For such formulas, the 3D view encodes every data element as absolute difference between its $R^2$ features and the global mean along the z-axis.
%%
This highlights slices with unusually low or high results (Fig.~\ref{fig:Application}).
%%
\magenta{Variables are ordered the same way in the 2D mosaic plot and 3D cube to preserve the mental model and make them visually analogous.}

\paragraph{Tackling the disadvantages of 3D information visualization.}
%%
3D information visualizations are often criticized for introducing occlusions and interaction problems.
%%
These are often not balanced out by the advantages of using the third dimension for visual mapping.
%%
We aim to minimize these problems.
%%
Since the $R^2$ values are mapped on data point opacity, large values are highlighted in the prism, guiding the focus to the respective slices.
%%
It also creates a sparse representation, since the majority of regression models yield (depending on the data set and the chosen formula) low $R^2$ values.
%%
Also, the preceding correlation-based feature selection reduces the information space significantly, leading to sparse cubes.
%%
Overlapping is still an issue, but this way greatly reduced in its effect to the readability of the visualization.
%%

>>>>

%\subsection{Abstracting Regression Results Using $R^2$}
\subsection{Abstracting Regression Results}
\begin{figure}[htb]
 \centering
 \includegraphics[width=1.0\linewidth]{figures/cube}
 \caption{
 (a) Overview visualization using a mosaic plot for the formula $Z \sim X + Y$, where $Z$ assumes the feature $age$.
 %%
 The descriptive metrics extracted from the regression model are mapped to color saturation (for $R^2$, saturated color shows a strong correlation).
 %%
 (b) Now, $Z$ is set to all features $n$ and yields $n$ mosaic plot visualizations.
 %%
 These represent the slices in our cube.
 %%
 The metric describing the regression model of each slice voxel is mapped on opacity in the 3D view later on, reducing the occlusion of other values.
 }
  \label{fig:Cube}
\end{figure}
%%
\noindent The goal of an overview visualization is to provide a comprehensive view on the data (raw or using descriptive metrics \cite{Bertini}), which is easy to understand.
%%
As described in our previous work \cite{Klemm2014VIS}, correlation values scaled between 0 (no correlation) and 1 (perfect correlation) can be encoded with color in a mosaic plot.
%%
Regression models are more complex, having many associated describing metrics.
%%
For the \emph{Regression Cube} analysis we are interested in the quality-of-fit of the resulting model.
%%
This allows to infer about the predictive quality of the independent features included in the model.
%%
As described in Section~\ref{sec:RegressionAnalysis}, the $R^2$ \add{, adjusted $R^2$ and AIC} value are metrics allowing for this kind of assessment.
%%
% \add{
% %%
% Users with in-depth knowledge about regression models are given the option to render the \emph{Regression Cube} using other descriptive features presented in \ref{sec:RegressionAnalysis}.
% %%
% Therefore, these experts can investigate the cube with emphasis on specific aspects of the model.
% %%
% To avoid shutting out users unfamiliar with these metrics, the \emph{Regression Cube} is set per default to show $R^2$ values.
% }
%%
\paragraph{2D (Slice) View.}
%%
Since $R^2$ is scaled between $[0, 1]$, it allows for comparison \emph{between} regression models.
%%
We can apply the same mosaic plot mapping by translating the $R^2$ values to \emph{color saturation} (Fig.~\ref{fig:Cube}a).
%%
This describes a 2D regression square for dynamic variables $X$ and $Y$ (e.g., $Age \sim X + Y$).
%%
\design{
%%
Based on expert feedback on early versions of this view, the amount of features used to compare regression models was extended.
%%
Therefore, these experts can investigate the cube with emphasis on specific aspects of the model.
%%
}
%%
\add{
%%
$Adjusted$ $R^2$ can be represented in the same way, since they share the scale.
%%
$AIC$ values are not restricted to a normalized scale.
%%
They have to be normalized in order to map it onto color saturation.
%%
%The resulting scale may be distort by outliers, which requires facilities for allowing the user to select scale ranges.
The resulting scale may be distorted by outliers derived from poor regression models.
%%
To tackle this problem, we provide a slider input, which maps the transfer function of the metric to color saturation based on user-selected ranges.
%%
Outliers can be cut off to emphasize ranges of interest.
%%
}
\design{
%%
To avoid shutting out users unfamiliar with these metrics, the \emph{Regression Cube} is set per default to show $R^2$ values.
%%
}

@@

\textbf{Show the Important} [\textbf{2.SI}]. The 3D cube visualization acts as an overview over the whole data set.
%%
Here, regression models with large regression metric values can be spotted fast, steering the user's attention to the respective slice.

@@

$Cancer \sim X + Y + Z$ subtracted with the regression metric from $Cancer \sim Age$ excludes the confounding effect that age has in view of the target $Cancer$ feature.
%%
This is achieved through cube comparison.
%%

\paragraph{Cube Comparison.}
\design{
%%
Cube comparisons were introduced later in the project.
%%
Epidemiologists with focus on statistics pointed out, that comparing outcomes of different formulas is suitable for removing the effect of possible confounding features.
%%
}
%%
\emph{Regression Cubes} can be compared by creating difference views.
%%
One cube (extracted with one formula) acts as reference.
%%
The absolute difference in the regression metric values with the second cube is calculated, yielding a difference cube showing only the differences between the two formulas.
%%
For example, it can be utilized for comparing the influence of a single feature on the complete result (e.g., $Z \sim X + Y$ and $Z \sim X + Y + Income$).

@@

%%
\textbf{Difference Cube.}
Adding a formula also adds it to the reference selection for a difference cube.
%%
Since all cells in the cube are represented using regression metric values, a difference cube is calculated by denoting the absolute difference of regression metric for each cell.
%%
\subsection{Regression Cube Visualization [2.SI].}
%%
The visualization and interaction with the \emph{Regression Cube} is the prototype core.
%%
Results from the statistical processors are uploaded into the visualization slice by slice.
%%
This allows the assessment of the data as soon as parts of the calculations are finished while the rest is still in progress.

@@

\paragraph{3D Prism as Data Mini-Map.}
\design{
%%
In early prototype versions, the 3D prism acted as starting point for the data analysis without the implementation of a separate 2D view.
%%
Slices were shown using cutaway planes.
%%
This approach was not popular with our domain experts, because the complexity of the visualization overwhelmed them. %did not like the combination of 3D interaction and slicing with representatives.
%%
}
%%
The 3D cube representation \design{was redesigned to} act as an overview over the whole data set.
%%
Its purpose is not to derive detailed information about data points.
%%
It serves as a function similar to a mini-map, guiding the attention to points of interest in the data.
%%
It also gives context information about adjacent data values when using the 2D mosaic plot.
%%
\design{The distinction between overview and details and demand using two different representations was well received with our domain experts.}
%%
The displayed prism shows values above the matrix diagonal.
%%
For formulas with a dynamic target feature (e.g. exploratory analysis using $Z \sim X + Y$), the color encodes the absolute regression metric values (Fig.~\ref{fig:System}b).
%%
Applying this strategy to a formula containing a static target (e.g. $Cancer \sim X + Y + Z$) yields many occlusions, since the CFS algorithm creates the same feature space for every slice.
%%
For such formulas, the 3D view encodes every data element as absolute difference between its regression metric values and the global mean along the z-axis.
%%
This highlights slices with unusually low or high results (Fig.~\ref{fig:Application}).
%%
\magenta{Variables are ordered the same way in the 2D mosaic plot and 3D cube to preserve the mental model and make them visually analogous.}

\paragraph{Tackling the disadvantages of 3D information visualization.}
%%
3D information visualizations are often criticized for introducing occlusions and interaction problems.
%%
These are often not balanced out by the advantages of using the third dimension for visual mapping.
%%
We aim to minimize these problems.
%%
Since the regression metric (e.g. $R^2$) values are mapped on data point opacity, large values are highlighted in the prism, guiding the focus to the respective slices.
%%
It also creates a sparse representation, since the majority of regression models yield (depending on the data set and the chosen formula) low $R^2$ values.
%%
Also, the preceding correlation-based feature selection reduces the information space significantly, leading to sparse cubes.
%%
Overlapping is still an issue, but this way greatly reduced in its effect to the readability of the visualization.
%%
